---
title: "Week 34 Mixed ANOVA"
author: "Dinkar Sharma, Lazaros Gonidis"
date: "21/03/2022"
output:
  html_document: default
---

```{r setup, include=FALSE}
options(digits = 3)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Simple effects analysis is relatively straightforward if you can carry out simple effects analysis that involve only within-participants factors.  In such cases you do not use the ‘pooled’ error terms but carry out the smaller repeated measures ANOVAs as you were shown in Exercise on repeated measures ANOVA. In other words, break up the between-participants factor(s) and analyse whatever is left over. For example, consider a 2x2x2 design, with three factors called A, B, and C, where Factor A is between-participants and B and C are within-participants. If the three-way interaction is significant, you could split the file by Factor A, and look at the simple interaction between B and C in condition 1 of A, and then the simple interaction between B and C in condition 2 of A. Because you are left only analysing within-participants factors, you can trust the R output and do not have to use a pooled error term.    


## Today's exercises   

In today's tasks you are going to use the data file called **mixed_anova.csv**. This data file includes the stress levels of patients and spouses (both male and female) at time 1, time 2 and time 3. This exercise involves a mixed-model ANOVA with three factors (a 3 x 2 x 2 design). The first factor of **time** has three levels which represent the stress levels being repeatedly measured at time 1 (the variable stress1), 2 (the variable stress2) and 3 (the variable stress3). The second factor in the design (**gender**) has two levels and is a between subjects factor. If the respondent is male then gender = 1; if female, gender = 2. The third factor (**role**) in the design has two levels and is a between subjects factor.  If the respondent is a patient then role = 1; if a spouse of the patient, role = 2. 

Factor 1: time, three levels, repeated measures.   
Level 1 = stress1   
Level 2 = stress2   
Level 3 = stress3    
  
Factor 2: gender, two levels, between subjects.       
Level 1 = male    
Level 2 = female    

Factor 3: role, two levels, between subjects.    
Level 1 = patient   
Level 2 = spouse    

DV: score, reported stress

Some of the packages you might need this week.   

`library(tidyverse) ### for general purpose functions`   
`library(rstatix) ### used for our ANOVA, anova_test()`
`library(ggpubr) ### used for graphs such as gghistogram()`     
  

##Task 1:  Mixed Model Three factor ANOVA.   

Carry out the Mixed model ANOVA.   


Q1) Write a summary table of the analysis ensuring all 7 effects are presented. Include appropriate error terms.   

Q2) What are the mean stress levels for male patients at time 1, time 2 and time 3?   

Q3) What are the mean stress levels for female spouses at time 1, time 2 and time 3?   

Q4) What is the marginal mean stress level for male spouses?    

Q5) What is the marginal mean stress level for female patients?   

Q6) Plot the three-way interaction by role and gender for each level of time? What interpretation would you offer for it?    

Q7) Given that the three-way interaction effect is significant, are the other significant effects of interpretative importance?    
 

## Task 1: Solution 

```{r task1, echo=TRUE, message=FALSE, warning=FALSE}
library(haven) ### used to load SPSS .sav files, read_sav()
library(tidyverse) ### for general purpose functions
library(rstatix) ### used for our ANOVA, anova_test()
library(ggpubr) ### used for graphs such as gghistogram()
library(emmeans)

df <- read_csv("mixed_anova.csv")


df$gender <- factor(df$gender, levels = c(1,2),
                    labels = c("male", "female"))

df$role <- factor(df$role, levels = c(1,2),
                     labels = c("patient", "spouse"))

df$fno <- as.factor(df$fno)

df$time <- as.factor(df$time)
  
  

  
anova1 <- anova_test(data =df, dv = score, 
                     wid = fno,
                     between = c(gender, role),
                     within = time,
                     type = 3,
                     detailed = TRUE)

anova1


df %>% 
  group_by(gender, role, time) %>% 
  summarise(mean = mean(score))


df %>% 
  group_by(gender, role) %>% 
  summarise(mean = mean(score))



ggline(df, x = "time", 
       y = "score", 
       color = "role",
       facet.by = "gender",
       add = "mean_se"
       )
```





## Task 2 : Simple effects analysis for a three way mixed model ANOVA.

	In view of the significant three-way interaction in the results of task 1, simple effects analyses need to be carried out. A useful simple effect analysis for this design would be to separate each cell of the 2 x 2 between subject factor and analyse the effects of stress levels. These are the simple main effects analysis we are going to carry out here. Carry out each of the following simple main effects analyses:

2(a) Test the difference in stress levels at time 1, 2 and 3 with male patients..
2(b): Test the difference in stress levels at time 1, 2 and 3 with male spouses
2(c): Test the difference in stress levels at time 1, 2 and 3 with female patients
2(d): Test the difference in stress levels at time 1, 2 and 3 with female spouses

Q1) Report the F-value, degrees of freedom, significance level and MS error (error variance) for each tested simple main effect..
Q2) Give a verbal interpretation of each tested simple effect.
Q3) Given the pattern of simple effects, what interpretation would you offer for the significant three-way interaction obtained in the ANOVA? (Hint: plot a graph of the two-way interaction within each level of time)
  



## Task 2: Solution   


```{r task2.solution3, echo=TRUE, message=FALSE, warning=FALSE, error=TRUE}

df %>% 
  group_by(gender, role) %>% 
  anova_test(dv = score, 
             wid = fno,
             within = time,
             type = 3,
             detailed = TRUE) %>% 
  get_anova_table()

ggline(df, x = "gender", 
       y = "score", 
       color = "role",
       facet.by = "time",
       add = "mean_se"
       )

```
