---
title: "Week 26. T-test Exercise"
author: "Dinkar Sharma, Lazaros Gonidis"
date: "18/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In today's session we want you to revisit the *t*-test methods you used in your Stage 1. Use the **Stress.csv** data file and the appropriate arguments for the **t_test()** function. Before doing that have a look at the brief refresher below on how to check the t-test assumptions and we carry out the appropriate t-test.


## T-test preliminary steps and assumptions

1. Normality assumption  
  1.1 Shapiro test  
  1.2 qq-plot  
  1.3 Skewness and Kurtosis  
2. Equal variance assumption  
  2.1 Calculate variance or standard deviation  
  2.2 Levene's test  
3. Carry out the appropriate t-test

## Let us have a look at how we can carry our these checks using the datafile **student t test.csv**. 

```{r example, echo=TRUE, message=FALSE, warning=FALSE}
library(rstatix)
library(tidyverse)
library(ggpubr)

### load file
df <- read_csv("student t test.csv")

### group as factor
df$group <- as.factor(df$group)

### descriptives
group_by(df, group) %>% 
  get_summary_stats(performance, type = "mean_sd") %>% 
  ungroup()

### check whether our scores are normally distributed
### using Shapiro Wilk test
### p values above 0.05 indicate normally distributed scores
df %>%
  group_by(group) %>%
  shapiro_test(performance) %>% ### Checks for normality
  ungroup()

### For Shapiro Wilk test we want to see p values above 0.05
### if both p values come back above 0.05 then you can proceed


### we can also inspect for normality using a qq plot
### this is a plot that show how our scores are correlated
### to a normal distribution
### visually if all our scores fall into the grey area indicate
### that they are normally distributed

ggqqplot(df, x = "performance", facet.by = "group")


### we can also get skewness and kurtosis information
### for that we need to have installed the moments package

### make sure  you install the package moments before proceeding

### you can do so by running install.packages("moments") 
### you should run this line only once
library(moments)

df %>%
  group_by(group) %>%
  summarise(Skew = skewness(performance),
            Kurt = kurtosis(performance))


### check for equal variances
### p value above 0.05 indicate no violation of the assumption
### of equal variance

levene_test(df, performance ~ group)  ### performance by group

### If Levene's test returns a p-value above 0.05 then my assumption
### of equal variance is not violated.
### time for the t-test statistic calculation
### var.equal = TRUE declares that we have equal variance
### therefore we can use Student's t-test
### var.equal = FALSE declares that we have unequal variance
### therefore we should use Welch's t-test

```



By looking at our output we see that the two groups are normally distributed and have equal variance. We will therefore run **Student's t-test**.

```{r example student's}
t_test(df, performance~group, var.equal = TRUE)
```







# Task 1

Here we are going to test whether or not there is a difference in the overall stress levels at time one and time two.  Before doing this, you must obtain descriptive statistics (Mean, Standard deviation, Variance, Kurtosis and Skewness) for STRESS1 and STRESS2 and produce a histogram for each of the variables. Once these statistics and histograms have been generated, carry out the repeated measures *t*-test.


Q1) Will you be using a paired-samples or an independent-samples t-test?

Q2) Which argument did you use for running the correct t-test?

Q3) What are the assumptions of this test and judging from the histograms and variances, to what extent are these assumptions violated by the data? 

Q4) What is the Null Hypothesis for this test?

Q5) Can it be rejected?________. Fill in :  t (     ) = ______, ______

